---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
```
# Irish Energy Balance Statistics

<!-- badges: start -->
<!-- badges: end -->

```{r packages, message=FALSE}
library("pxR")
library("tidyverse")
library("janitor")
library("lubridate")
library("hrbrthemes")
library("ggstance")
library("gganimate")
```

_Disclaimer: I am not an expert in energy. I quickly  realised the following visualisations for my first TidyTuesday, and to learn how to use {[gganimate](https://github.com/thomasp85/gganimate)}. I'm sure I made some mistakes along the way and that things could look a little bit better._

## Data

Source: [Sustainable Energy Authority of Ireland](https://www.cso.ie/px/pxeirestat/Database/eirestat/Energy%20Balance%20Statistics/Energy%20Balance%20Statistics_statbank.asp?SP=Energy%20Balance%20Statistics&Planguage=0&ProductID=DB_SEI).


```{r}
energy_balance_px <- read.px("data/SEI01.px")
energy_balance_tbl_raw <- as.data.frame(energy_balance_px) %>%
  as_tibble() %>%
  clean_names()

sample_n(energy_balance_tbl_raw, 10)
```


## Munging

```{r}
energy_balance_tbl <- energy_balance_tbl_raw %>%
  mutate_if(is.factor, as.character) %>% 
  mutate_at(vars(year), parse_date_time, orders = "%Y") %>%
  filter(!str_detect(fuel_type, "^Sum of all")) %>%
  rename(energy = value) %>%
  filter(!fuel_type %in% c("Heat", "Non renewable waste")) %>% # remove fuels I don't understand
  mutate(
    fuel_type = str_replace(fuel_type, "Electricity", "imported electricity"),
    fuel_type_category = fct_collapse(fuel_type,
      renewable = c("Hydro", "Wind", "Landfill gas", "Biogas", "Biomass", "Liquid Biofuel", "Solar", "Geothermal"),
      petroleum = c("Crude oil", "Refinery gas", "Gasoline", "Kerosene", "Jet kerosene", "Fuel oil", "LPG", "Gasoil/diesel/DERV", "Petroleum coke", "Naphta", "Bitumen", "White spirit", "Lubricants", "Natural gas"),
      coal = c("Coke", "Lignite", "Bituminous coal", "Anthracite and manufactured ovoids"),
      peat = c("Milled peat", "Sod peat", "Briquettes")
    )
  ) %>%
  filter(!str_detect(energy_supply_and_consumption, "[Ss]um of all")) %>% # remove summary rows
  filter(!energy_supply_and_consumption %in% c("Total final energy consumption", "Final non energy consumption feedstocks", "Available final energy consumption", "Statistical difference", "Primary energy supply including non energy", "Primary energy requirement excluding non energy")) %>%
  mutate(
    energy_supply_and_consumption_subsubcategory = fct_collapse(energy_supply_and_consumption, # heavy usage of datapasta
      `indigenous production` = "Indigenous production",
      imports = "Imports",
      exports = "Exports",
      `stock change` = "Stock change",
      bunkers = "Mar. bunkers",
      `energy distribution` = c("Own use and distribution losses"),
      residential = c("Residential"),
      agricultural = c("Agricultural"),
      fisheries = c("Fisheries"),
      `commercial and public services` = c("Commercial services", "Public services"),
      aviation = c("Transport- fuel tourism", "Transport- international aviation", "Transport- domestic aviation"),
      road = c("Transport- road freight", "Transport- road private car", "Transport- road public passenger services", "Transport- road light goods vehicle"),
      rail = c("Transport- rail"),
      navigation = c("Transport - navigation"),
      `other transport` = "Transport- unspecified",
      industry = c("Industry- non energy mining", "Industry- food, beverages and tobacco", "Industry- textiles and textile products", "Industry- wood and wood products", "Industry- pulp, paper, publishing and printing", "Industry- chemicals and man-made fibres", "Industry- rubber and plastic products", "Industry- other non-metallic mineral products", "Industry- basic metals and fabricated metal products", "Industry- machinery and equipment n.e.c", "Industry- electrical and optical equipment", "Industry- transport equipment manufacture", "Industry- other manufacturing"),
      `transformation input` = c("Transformation input- public thermal power plants", "Transformation input- combined heat and power plants", "Transformation input- pumped storage consumption", "Transformation input- briquetting plants", "Transformation input- oil refineries and other energy sectors"),
      `transformation output` = c("Transformation output- public thermal power plants", "Transformation output- combined heat and power plants- electricity", "Transformation output- combined heat and power plants- heat", "Transformation output- pumped storage consumption", "Transformation output- briquetting plants", "Transformation output- oil refineries"),
      `echanges and transfer` = c("Exchanges and transfers- electricity", "Exchanges and transfers- heat", "Exchanges and transfers- other (excluding electricity and heat)"),
      `non-energy consumption` = "Non-energy consumption"
    ),
    energy_supply_and_consumption_subcategory = fct_collapse(energy_supply_and_consumption_subsubcategory,
      transport = c("navigation", "rail", "road", "aviation", "other transport"),
      rci = c("residential", "commercial and public services", "industry", "agricultural", "fisheries"),
    ),
    energy_supply_and_consumption_category = fct_collapse(energy_supply_and_consumption_subcategory,
      consumption = c("transport", "rci"),
      `supply gains` = c("indigenous production", "imports"),
      `supply loses` = c("exports", "bunkers")
    )
  ) %>%
  mutate_at(vars(energy_supply_and_consumption), str_remove, pattern = "^.*-")
```

## Visualisations


### Renewable energy

```{r fig.height=5.5, fig.width=4.5}
renewable_plot <- energy_balance_tbl %>%
  filter(energy_supply_and_consumption_category == "supply gains") %>%
  filter(!fuel_type == "Electricity") %>%
  mutate(is_renewable = fuel_type_category == "renewable") %>%
  group_by(year, is_renewable) %>%
  summarise(energy = sum(energy)) %>%
  ggplot(mapping = aes(x = year(year), y = energy, fill = is_renewable, colour = is_renewable)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_hline(yintercept = 0.16 - 0.01, size = 0.5, linetype = "dotted") +
  annotate("text", x = 1990, y = 0.16, label = "2020 target", colour = "black", hjust = 0, vjust = 0, alpha = 0.8) +
  scale_y_continuous(labels = scales::percent, sec.axis = dup_axis(breaks = c(0.0863078, 0.16), labels = scales::percent)) +
  scale_x_continuous(breaks = c(1990, 2000, 2010, 2017)) +
  scale_color_ipsum() +
  scale_fill_ipsum() +
  theme_ipsum_pub() +
  theme(
    legend.position = "none",
    axis.text = element_text(),
    axis.title.y.right = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Renewable energy",
    subtitle = "in Republic Of Ireland, 1990-2017",
    x = "year",
    y = "% energy supply",
    caption = "data: SEAI/CSO"
  )

renewable_plot

ggsave("renewable_energy_roi.png")
```




Ireland is only green from the sky. More renewable source of energies would also make the island less dependent from other countries petrol.

### Energy production 

```{r}
production_anim <- energy_balance_tbl %>%
  filter(energy_supply_and_consumption_category == "supply gains") %>%
  mutate(fuel_type = fct_reorder(fuel_type, as.numeric(fuel_type_category))) %>%
  ggplot(mapping = aes(x = energy, y = fuel_type, fill = fuel_type_category)) +
  geom_barh(stat = "identity") +
  theme_ipsum_pub() +
  scale_fill_ipsum(guide = guide_legend(nrow = 2)) +
  theme(
    legend.position = c(1, 1),
    legend.justification = c(1, 1),
    legend.direction = "horizontal",
    axis.title.y = element_blank()
  ) +
  labs(
    title = "Source of energy",
    subtitle = "in Republic of Ireland, year: {year(frame_time)}",
    x = "energy (kilotonne of oil equivalent)",
    fill = "",
    caption = "data: SEAI/CSO"
  ) +
  transition_time(year) +
  enter_grow()

animate(production_anim, width = 600, height = 600, units = "px")

anim_save("production_energy_roi.gif")
```

### Consuption


```{r}
consuption_anim <- energy_balance_tbl %>%
  filter(energy_supply_and_consumption_category == "consumption") %>% 
  group_by(year, energy_supply_and_consumption_subsubcategory, energy_supply_and_consumption_subcategory) %>% 
  summarise(energy = sum(energy)) %>% 
  ggplot(mapping = aes(x = energy, y = energy_supply_and_consumption_subsubcategory, fill = energy_supply_and_consumption_subcategory)) +
  geom_barh(stat = "identity") +
  theme_ipsum_pub() +
  scale_fill_brewer(palette = "Set2", guide = guide_legend(nrow = 2)) +
  theme(
    legend.position = c(1, 1.15),
    legend.justification = c(1, 1),
    legend.direction = "horizontal",
    axis.title.y = element_blank()
  ) +
  labs(
    title = "Consuption of energy",
    subtitle = "in Republic of Ireland, year: {year(frame_time)}",
    x = "energy (kilotonne of oil equivalent)",
    fill = "",
    caption = "data: SEAI/CSO"
  ) +
  transition_time(year) +
  enter_grow()
  

animate(consuption_anim, width = 640, height = 500, units = "px")

anim_save("consuption_energy_roi.gif")
```



## Session info

```{r fig.width=10}
sessioninfo::session_info()
```

